Analysis on Trend of Death of Shelter Residence in Toronto 

### 2025-06-21 paper/paper.qmd
changed paper scripts

```{r}
#| include: false
#| warning: false
#| message: false

library(kableExtra)
library(tidyverse)
library(palmerpenguins)

# Read the analysis data of shelter residence death.
shelter_death_data <-
  read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv",
    show_col_types = FALSE
  )
```

@@

### Trend Over Time {#sec-time}
@fig-shelter-death-line, created by @citeR, shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year. This layout allows readers to clearly compare the trend of deaths from one year to the next and observe potential seasonal changes.

```{r fig-shelter-death-line, fig.cap="Deaths in Toronto Shelters Over Time (2007.01-2024.08)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
# Sort the month data in order
shelter_death_data$month <- factor(shelter_death_data$month, levels = month.abb)

# Get line graph, showing the death trend
ggplot(shelter_death_data, aes(x = month, y = total_decedents, group = year)) +
  geom_line(size = 1, color = "black") + # get line graph
  geom_point(size = 2, color = "pink") + # add points on value in each year
  labs(
    title = "Deaths in Toronto Shelters Over Time (2007.01-2024.08)",
    x = "Month",
    y = "Total Deaths"
  ) + # add title and name of the axis
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
    # Rotate horizontal axis labels to avoid overlapping
  ) +
  facet_wrap(~year, ncol = 3)
```
Based on the generated line graphs, the following conclusions can be drawn:

@@

```{r fig-shelter-death-bar, fig.cap="Annual Death Counts in Toronto Shelters With Gender Distribution (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
annual_death_counts_bar <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2023) %>%
  group_by(year) %>%
  summarise(
    total_deaths = sum(total_decedents),
    male_deaths = sum(male),
    female_deaths = sum(female),
    other_deaths = sum(transgender_non_binary_two_spirit, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(male_deaths, female_deaths, other_deaths),
    names_to = "gender",
    values_to = "count"
  )

# Create bar chart showing annual death counts in Toronto shelters
# from 2007 to 2023.
ggplot(annual_death_counts_bar, aes(
  x = factor(year),
  y = count,
  fill = gender
)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Annual Death Counts in Toronto
       Shelters With Gender Distribution (2007-2023)",
    x = "year",
    y = "num of death",
    fill = "gender"
  ) +
  scale_fill_manual(
    values = c(
      "male_deaths" = "pink",
      "female_deaths" = "black",
      "other_deaths" = "gray"
    ),
    labels = c("Female", "male", "Other")
  ) +
  # change color and the text on the side
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
    # Rotate horizontal axis labels to avoid overlapping
  )
```

```{r fig-annual-death, fig.cap="Annual Deaths in Toronto Shelters (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
# calculate the annual death
annual_deaths <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2023) %>%
  group_by(year) %>%
  summarise(total_yearly_deaths = sum(total_decedents), na.rm = TRUE)

# get line graph, showing the annual death trend
ggplot(annual_deaths, aes(x = year, y = total_yearly_deaths)) +
  geom_line(size = 1, color = "black") + # get line graph
  geom_point(size = 3, color = "pink") + # add points on value in each year
  labs(
    title = "Annual Deaths in Toronto Shelters (2007-2023)",
    x = "Year",
    y = "Total Deaths"
  ) + # add title and name of the axis
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8)
  )
```

@@

```{r fig-monthly-death, fig.cap="Monthly Mortality Trends Among Shelter Residents (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
# calculate the annual death
month_deaths <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2023) %>%
  group_by(month) %>%
  summarise(total_monthly_deaths = sum(total_decedents))

month_deaths$month <- factor(month_deaths$month, levels = month.abb)

# get line graph, showing the annual death trend
ggplot(month_deaths, aes(x = month, y = total_monthly_deaths, group = 1)) +
  geom_line(size = 1, color = "black") + # get line graph
  geom_point(size = 3, color = "pink") + # add points on value in each year
  labs(
    title = "Monthly Mortality Trends Among Shelter Residents (2007-2023)",
    x = "Month",
    y = "Total Deaths"
  ) + # as title and name of the axis
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8)
  )
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get first 20 lines in the data set as a sample
example_table <- shelter_death_data %>%
  slice(1:5)

example_table %>%
  kbl(
    caption = "Sample Table of Deaths in Toronto Shelters (2007-2023)",
    col.names = c(
      "ID", "Year", "Month", "Total decedents",
      "male", "female", "Others"
    )
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    latex_options = "scale_down"
  ) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# summarize the data group by year
summary_table <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2023) %>%
  group_by(year) %>%
  summarise(
    total_deaths = sum(total_decedents), # total death num
    mean_deaths = round(mean(total_decedents), 2),
    # average death num each month, rounded to 2 decimal places
    median_deaths = round(median(total_decedents), 2),
    # median, rounded
    min_deaths = min(total_decedents), # minimum value
    max_deaths = max(total_decedents), # maximum value
    male_ratio = round(sum(male) / total_deaths * 100, 2),
    # ratio of male death, rounded
    female_ratio = round(sum(female) / total_deaths * 100, 2),
    # ratio of female death, rounded
    other_ratio =
      ifelse(any(is.na(shelter_death_data$transgender_non_binary_two_spirit)),
        NA,
        round(sum(transgender_non_binary_two_spirit,
          na.rm = TRUE
        ) / total_deaths * 100, 2)
      )
    # ratio of death of other genders, rounded
  )
summary_table %>%
  kbl(
    caption = "Annual Summary of Deaths in Toronto Shelters (2007-2023)",
    col.names = c(
      "Year", "Total Deaths", "Mean", "Median", "Min", "Max",
      "Male Ratio (%)", "Female Ratio (%)", "Other Ratio (%)"
    )
  ) %>%
  kable_styling(
    full_width = FALSE, position = "center",
    latex_options = "scale_down"
  ) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  add_header_above(c(" " = 1, "Deaths Summary" = 5, "Gender Ratios" = 3))
```

>>>>

```{python}
#| include: false
#| warning: false
#| message: false

import pandas as pd
import matplotlib.pyplot as plt
import numpy as np

# Read the analysis data of shelter residence deaths
shelter_death_data = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
```

@@

### Trend Over Time {#sec-time}
@fig-shelter-death-line, created by @citeR, shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year. This layout allows readers to clearly compare the trend of deaths from one year to the next and observe potential seasonal changes.
```{python}
#| label: fig-shelter-death-line
#| fig-cap: "Deaths in Toronto Shelters Over Time (2007.01–2024.08)"
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 6

# Load the data
df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")

# Filter for years 2007–2023
filtered = df[(df["year"] >= 2007) & (df["year"] <= 2023)].copy()

# Group by year and compute summary stats
summary = filtered.groupby("year").agg(
    total_deaths=("total_decedents", "sum"),
    mean_deaths=("total_decedents", lambda x: round(x.mean(), 2)),
    median_deaths=("total_decedents", lambda x: round(x.median(), 2)),
    min_deaths=("total_decedents", "min"),
    max_deaths=("total_decedents", "max"),
    male_sum=("male", "sum"),
    female_sum=("female", "sum"),
    other_sum=("transgendernonbinarytwospirit", lambda x: x.sum(skipna=True))
).reset_index()

# Calculate gender ratios
summary["male_ratio"] = round(summary["male_sum"] / summary["total_deaths"] * 100, 2)
summary["female_ratio"] = round(summary["female_sum"] / summary["total_deaths"] * 100, 2)

# Handle possible NA case in other gender ratio
if df["transgendernonbinarytwospirit"].isna().any():
    summary["other_ratio"] = np.nan
else:
    summary["other_ratio"] = round(summary["other_sum"] / summary["total_deaths"] * 100, 2)

# Keep only relevant columns and rename for display
summary_table = summary[[
    "year", "total_deaths", "mean_deaths", "median_deaths", "min_deaths", "max_deaths",
    "male_ratio", "female_ratio", "other_ratio"
]].rename(columns={
    "year": "Year",
    "total_deaths": "Total Deaths",
    "mean_deaths": "Mean",
    "median_deaths": "Median",
    "min_deaths": "Min",
    "max_deaths": "Max",
    "male_ratio": "Male Ratio (%)",
    "female_ratio": "Female Ratio (%)",
    "other_ratio": "Other Ratio (%)"
})

# Display styled table with caption
summary_table.style.set_caption("Annual Summary of Deaths in Toronto Shelters (2007–2023)") \
    .set_table_styles([
        {"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]},
        {"selector": "th", "props": [("text-align", "center")]},
        {"selector": "td", "props": [("text-align", "center")]}
    ]) \
    .set_properties(**{"font-family": "Cambria"}) \
    .format(na_rep="NA")
```
Based on the generated line graphs, the following conclusions can be drawn:

@@

```{python} 
#| label: fig-shelter-death-bar
#| fig-cap: "Annual Death Counts in Toronto Shelters With Gender Distribution (2007–2023)"
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

import pandas as pd
import matplotlib.pyplot as plt

# Load the data
df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")

# Filter for years 2007–2023
df = df[(df["year"] >= 2007) & (df["year"] <= 2023)]

# Group and sum by year
grouped = df.groupby("year").agg(
    male_deaths=("male", "sum"),
    female_deaths=("female", "sum"),
    other_deaths=("transgendernonbinarytwospirit", "sum")
).fillna(0)

# Convert to long format
long_df = grouped.reset_index().melt(id_vars="year", var_name="gender", value_name="count")

# Set color mapping and label override
color_map = {
    "male_deaths": "pink",
    "female_deaths": "black",
    "other_deaths": "gray"
}
label_map = {
    "male_deaths": "male",
    "female_deaths": "Female",
    "other_deaths": "Other"
}

# Pivot for stacked bar
pivot_df = long_df.pivot(index="year", columns="gender", values="count").fillna(0)
pivot_df = pivot_df[color_map.keys()]  # Ensure column order

# Plot
fig, ax = plt.subplots(figsize=(10, 6))

bottom = None
for gender in pivot_df.columns:
    ax.bar(pivot_df.index, pivot_df[gender], label=label_map[gender],
           bottom=bottom, color=color_map[gender])
    bottom = pivot_df[gender] if bottom is None else bottom + pivot_df[gender]

# Labels and title
ax.set_title("Annual Death Counts in Toronto Shelters With Gender Distribution (2007–2023)",
             fontsize=16, fontweight='bold', ha='center')
ax.set_xlabel("Year", fontsize=12)
ax.set_ylabel("Num of Death", fontsize=12)
ax.set_xticks(pivot_df.index)
ax.set_xticklabels(pivot_df.index, rotation=45)
ax.tick_params(axis='both', labelsize=10)

# Legend and theme
ax.legend(title="Gender", title_fontsize=11)
ax.grid(True, axis='y', linestyle='--', alpha=0.7)
ax.set_axisbelow(True)

plt.tight_layout()
plt.show()

```

```{python}
#| label: fig-annual-death
#| fig-cap: "Annual Deaths in Toronto Shelters (2007–2023)"
#| echo: false
#| message: false
#| warning: false

import pandas as pd
import matplotlib.pyplot as plt

# Load the data
df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")

# Filter years and calculate total annual deaths
annual_deaths = (
    df[(df["year"] >= 2007) & (df["year"] <= 2023)]
    .groupby("year", as_index=False)
    .agg(total_yearly_deaths=("total_decedents", "sum"))
)

# Plot line chart
fig, ax = plt.subplots(figsize=(7, 4))

ax.plot(annual_deaths["year"], annual_deaths["total_yearly_deaths"],
        color="black", linewidth=1)
ax.scatter(annual_deaths["year"], annual_deaths["total_yearly_deaths"],
           color="pink", s=40)

# Labels and title
ax.set_title("Annual Deaths in Toronto Shelters (2007–2023)",
             fontsize=12, fontweight='bold', ha='center')
ax.set_xlabel("Year", fontsize=10)
ax.set_ylabel("Total Deaths", fontsize=10)
ax.tick_params(axis='both', labelsize=8)

# Apply minimal-style grid
ax.grid(True, linestyle='--', linewidth=0.5, alpha=0.7)
ax.set_axisbelow(True)

plt.tight_layout()
plt.show()

```

@@

```{python}
#| label: fig-monthly-death
#| fig-cap: "Monthly Mortality Trends Among Shelter Residents (2007–2023)"
#| echo: false
#| message: false
#| warning: false

# Load the data
df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")

# Filter for relevant years and group by month
month_deaths = (
    df[(df["year"] >= 2007) & (df["year"] <= 2023)]
    .groupby("month", as_index=False)
    .agg(total_monthly_deaths=("total_decedents", "sum"))
)

# Ensure correct month order
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
month_deaths["month"] = pd.Categorical(month_deaths["month"],
                                       categories=month_order,
                                       ordered=True)
month_deaths = month_deaths.sort_values("month")

# Plot
fig, ax = plt.subplots(figsize=(7, 4))

ax.plot(month_deaths["month"], month_deaths["total_monthly_deaths"],
        color="black", linewidth=1)
ax.scatter(month_deaths["month"], month_deaths["total_monthly_deaths"],
           color="pink", s=40)

# Labels and theme
ax.set_title("Monthly Mortality Trends Among Shelter Residents (2007–2023)",
             fontsize=12, fontweight='bold', ha='center')
ax.set_xlabel("Month", fontsize=10)
ax.set_ylabel("Total Deaths", fontsize=10)
ax.tick_params(axis='both', labelsize=8)

ax.grid(True, linestyle='--', linewidth=0.5, alpha=0.7)
ax.set_axisbelow(True)

plt.tight_layout()
plt.show()

```

@@

```{python}
#| echo: false
#| message: false
#| warning: false

# Load data
df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")

# Take first 5 rows as sample
example_table = df.iloc[:5].copy()

# Rename columns for display
example_table.columns = [
    "ID", "Year", "Month", "Total decedents",
    "male", "female", "Others"
] + list(example_table.columns[7:])  # In case there are more columns

# Display the table (nicely rendered in Quarto)
example_table.style.set_caption("Sample Table of Deaths in Toronto Shelters (2007–2023)") \
    .set_table_styles(
        [{"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]}]
    ) \
    .set_properties(**{"text-align": "center", "font-family": "Cambria"})

```

```{python}
#| echo: false
#| message: false
#| warning: false

# Load the data
df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")

# Filter for years 2007–2023
filtered = df[(df["year"] >= 2007) & (df["year"] <= 2023)].copy()

# Group by year and compute summary stats
summary = filtered.groupby("year").agg(
    total_deaths=("total_decedents", "sum"),
    mean_deaths=("total_decedents", lambda x: round(x.mean(), 2)),
    median_deaths=("total_decedents", lambda x: round(x.median(), 2)),
    min_deaths=("total_decedents", "min"),
    max_deaths=("total_decedents", "max"),
    male_sum=("male", "sum"),
    female_sum=("female", "sum"),
    other_sum=("transgendernonbinarytwospirit", lambda x: x.sum(skipna=True))
).reset_index()

# Calculate gender ratios
summary["male_ratio"] = round(summary["male_sum"] / summary["total_deaths"] * 100, 2)
summary["female_ratio"] = round(summary["female_sum"] / summary["total_deaths"] * 100, 2)

# Handle possible NA case in other gender ratio
if df["transgendernonbinarytwospirit"].isna().any():
    summary["other_ratio"] = np.nan
else:
    summary["other_ratio"] = round(summary["other_sum"] / summary["total_deaths"] * 100, 2)

# Keep only relevant columns and rename for display
summary_table = summary[[
    "year", "total_deaths", "mean_deaths", "median_deaths", "min_deaths", "max_deaths",
    "male_ratio", "female_ratio", "other_ratio"
]].rename(columns={
    "year": "Year",
    "total_deaths": "Total Deaths",
    "mean_deaths": "Mean",
    "median_deaths": "Median",
    "min_deaths": "Min",
    "max_deaths": "Max",
    "male_ratio": "Male Ratio (%)",
    "female_ratio": "Female Ratio (%)",
    "other_ratio": "Other Ratio (%)"
})

# Display styled table with caption
summary_table.style.set_caption("Annual Summary of Deaths in Toronto Shelters (2007–2023)") \
    .set_table_styles([
        {"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]},
        {"selector": "th", "props": [("text-align", "center")]},
        {"selector": "td", "props": [("text-align", "center")]}
    ]) \
    .set_properties(**{"font-family": "Cambria"}) \
    .format(na_rep="NA")
```
